Species richness and food-web structure jointly drive community biomass and its temporal stability in fish communities

### 2021-07-12 manuscript/bef_stability/manuscript.Rmd
According to comments of coauthors

Biodiversity-ecosystem functioning (BEF) and food-web complexity-stability
relationships are central to ecology. However, they remain largely untested in
natural contexts. Here we estimated the links among environmental conditions,
richness, food-web structure, annual biomass and its temporal stability using a
standardised monitoring dataset of 99 stream fish communities spanning from 1995
to 2018. We first revealed that both richness and average trophic level are
positively related to annual biomass, with effects of similar strength. Second,
we found that community stability is fostered by mean trophic level, while
contrary to expectation, it is decreased by species richness. Finally, we found
that environmental conditions affect both biomass and its stability mainly via
effects on richness and network structure. Strikingly,
community stability was mediated by the variation of population stability rather
than synchrony, which contrasts with results from single-trophic communities. We
discuss the hypothesis that it could be a characteristic of multi-trophic
communities.

@@

Environment can affect functioning and stability directly or through effects on
resource availability, species richness or food-web
structure. Agricultural intensification and temperature changes can also...

@@

The structural equation models explained a large part of the variation of
species richness, connectance, synchrony, $\overline{CV_{sp}}$ and community
biomass (resp. $R^2 = .43$, $.38$, $.46$, $.47$ and $.32$) but less so for mean
trophic level ($R^2 = .11$, Fig. \@ref(fig:sem-bm)A and \@ref(fig:sem-stab)A).
We found a negative relationship between species richness and mean trophic level
($r_\delta = `r rich_on_tlvl`$, $r_\delta$ meaning standardised slope) and
connectance ($r_\delta = `r rich_on_ct`$, Fig. \@ref(fig:sem-bm)A and
\@ref(fig:sem-stab)A), meaning that a part of the effects of species richness on
community biomass and its temporal stability went through food-web structure.

@@

Species richness and mean trophic level had a significant positive direct effect
on community biomass ($r_\delta = `r rich_on_bm$direct`$ and $r_\delta = 
`r tlvl_on_bm`$ resp., Fig. \@ref(fig:sem-bm)A, B, C, Table
\@ref(tab:indir-sem)),
while connectance had no direct effect.  The two PCA axes synthesising
environmental variables had no significant direct effects on community biomass,
but indirect ones (Table \@ref(tab:indir-sem)).  The first axis, related to the
stream size, was positively linked to community biomass ($r_\delta = 
`r env_on_bm$total["log_RC1"]`$, Table \@ref(tab:indir-sem)), through both
species
richness and mean trophic level 
($r_\delta = `r env_on_bm$via_richness["log_RC1"]`$ and 
$`r env_on_bm$via_tlvl["log_RC1"]`$
resp., Fig. \@ref(fig:sem-bm) F, G, Table \@ref(tab:indir-sem)). The second
axis, related to the average annual temperature and BOD was positively related
to community biomass ($r_\delta = `r env_on_bm$total["log_RC2"]`$, Table
\@ref(tab:indir-sem)) mainly through species richness ($r_\delta = 
`r env_on_bm$via_richness["log_RC2"]`$, Fig. \@ref(fig:sem-bm) G, Table
\@ref(tab:indir-sem)). Warmer and enriched, i.e.  with higher BOD, streams
contained more species, and thus indirectly increased the biomass of the
communities. The total effect of PCA1, PCA2 and species richness on community
biomass were of the same magnitude.

@@

In our study system, the destabilising effect of species richness via increased
population variability was stronger than the stabilising effect via decreased
synchrony.  This contrasts with the previous study of Olivier et al. (2021) on
animal communities in terrestrial ecosystems. The difference of the effect of
species richness on population stability and synchrony between studies might be
related to the multitrophic structure of the dataset as theory predicts
potential synchronising effect of generalist predators, such as piscivorous
fishes, on prey dynamics [@raimondo_interspecific_2004; @huber_role_2006].
Trophic interactions might also raise synchrony between predators and preys and
thereby increase the overall synchrony of trophic communities [Fig. S9,
@bulmer_phase_1975]. However, generalist predators can also generate
asynchronous dynamics among preys depending on the distribution of interaction
strengths [@huber_role_2006; @mccann_more_2009]. The difference between our
results and those of @olivier_urbanization_2020 might also relate to the type of
ecosystems considered. Aquatic ecosystems have been suggested to show greater
population variability than terrestrial ecosystems [@rip_cross-ecosystem_2011],
which could also translate in greater variation range of population variability
among aquatic communities than among terrestrial ones. Another hypothesis
relates to the evenness of the biomass distribution among species, with low
evenness decreasing the range of variation in synchrony among communities and
thereby the strength of the portfolio effect [@doak_statistical_1998;
@thibaut_understanding_2013]. Whether the biomass distribution in stream fish
assemblages is particularly uneven when compared to the terrestrial communities
studied by @olivier_urbanization_2020 remains however to be tested. In any case,
more studies are needed to understand how the relative importance of the effect
of species richness on synchrony and population variability varies among
ecosystems.

@@

sem_bm_caption <- paste0( 
  "Relationships among the predictors of community biomass. (A) Structural
  Equation Model for the median of annual community biomass (in grams by square
    metre). Dotted and solid arrows represent, respectively, negative and
  positive path coefficients. Non significant paths are coloured in grey. The
  path coefficients are standardised. $R^2$: adjusted $R^2$; Avg: Average, CVsp: weighted
  temporal coefficient of variation of the population biomass; PCA1 and PCA2:
  respectively the first and second axis of the principal component analysis
  performed on environmental variables (see methods). (B) to (G): significant
  bivariate relationships between variables of the structural equation model.
  Solid lines depict the predictions of the statistical model. A sensibility
  analysis is displayed in supplementary information (Fig.
    S8).")

@@

sem_stab_caption <- paste0(
  "Relationships among the predictors of the temporal stability of community
  biomass. (A) Structural Equation Model for the temporal stability of community
  biomass. Dotted and solid arrows represent, respectively, negative and positive
  path coefficients. Non significant paths are coloured in grey. The path
  coefficients are standardised. $R^2$: adjusted $R^2$; Avg: Average, CVsp: weighted
  temporal coefficient of variation of the population biomass; PCA1 and PCA2:
  respectively the first and second axis of the principal component analysis
  performed on environmental variables (see methods). The $R^2$ for the temporal
  stability of community biomass is exactly one because the relation is
  determistic (see Eq. 4). (B) to (I): significant
  bivariate relationships between variables of the structural equation model.
  Solid lines depict the predictions of the statistical model. A sensibility
  analysis is displayed in supplementary information (Fig.
    S8)."
  )
```

>>>>

Biodiversity-ecosystem functioning (BEF) and food-web complexity-stability
relationships are central to ecology. However, they remain largely untested in
natural contexts. Here we estimated the links among environmental conditions,
richness, food-web structure, annual biomass and its temporal stability using a
standardised monitoring dataset of 99 stream fish communities spanning from 1995
to 2018. We first revealed that both richness and average trophic level are
positively related to annual biomass, with effects of similar strength. Second,
we found that community stability is fostered by mean trophic level, while
contrary to expectation, it is decreased by species richness. Finally, we found
that environmental conditions affect both biomass and its stability mainly via
effects on richness and network structure. Strikingly,
the effect of species richness on community stability was mediated by population stability rather
than synchrony, which contrasts with results from single-trophic communities. We
discuss the hypothesis that it could be a characteristic of multi-trophic
communities.

@@

@@

The structural equation models explained a large part of the variation of
species richness, connectance, synchrony, $\overline{CV_{sp}}$ and community
biomass (resp. $R^2 = .43$, $.38$, $.46$, $.47$ and $.32$) but less so for mean
trophic level ($R^2 = .11$, Fig. \@ref(fig:sem-bm)A and \@ref(fig:sem-stab)A).
We found a negative relationship between species richness and mean trophic level
($r_\delta = `r rich_on_tlvl`$, $r_\delta$ meaning standardised slope) and
connectance ($r_\delta = `r rich_on_ct`$, Fig. \@ref(fig:sem-bm)A and
\@ref(fig:sem-stab)A), meaning that a part of the effects of species richness on
community biomass and its temporal stability went through food-web structure.
The relationships resulting from the SEM are robust to different station
selection criteria in the analysis, with different selections ranging in 65 to
403 stations (see sensitivity analysis, Fig. S8).

@@

Species richness and mean trophic level had a significant positive direct effect
on community biomass ($r_\delta = `r rich_on_bm$direct`$ and $r_\delta = 
`r tlvl_on_bm`$ resp., Fig. \@ref(fig:sem-bm)A, B, C, Table
\@ref(tab:indir-sem)),
while connectance had no direct effect.  The two PCA axes synthesising
environmental variables had no significant direct effects on community biomass,
but indirect ones (Table \@ref(tab:indir-sem)).  The first axis, related to the
stream size, was positively linked to community biomass ($r_\delta = 
`r env_on_bm$total["log_RC1"]`$, Table \@ref(tab:indir-sem)), through both
species
richness and mean trophic level 
($r_\delta = `r env_on_bm$via_richness["log_RC1"]`$ and 
$`r env_on_bm$via_tlvl["log_RC1"]`$
resp., Fig. \@ref(fig:sem-bm)F, G, Table \@ref(tab:indir-sem)). The second
axis, related to the average annual temperature and BOD was positively related
to community biomass ($r_\delta = `r env_on_bm$total["log_RC2"]`$, Table
\@ref(tab:indir-sem)) mainly through species richness ($r_\delta = 
`r env_on_bm$via_richness["log_RC2"]`$, Fig. \@ref(fig:sem-bm) G, Table
\@ref(tab:indir-sem)). Warmer and enriched, i.e.  with higher BOD, streams
contained more species, and thus indirectly increased the biomass of the
communities. The total effect of PCA1, PCA2 and species richness on community
biomass were of the same magnitude.

@@

In our study system, the destabilising effect of species richness via increased
population variability was stronger than the stabilising effect via decreased
synchrony. Further, as the effects of species richness on population variability
and synchrony we found here were of similar amplitude, the balance between both
pathways is governed by the relative standard deviation of population
variability and synchrony, this because we used standardized coefficients (see
methods). The fish communities studied here therefor present similar level of
variation in population variability and synchrony, again contrasting with
results on terrestrial animal communities that exhibited twice as much variation
in population synchrony than in population variability
[@olivier_urbanization_2020]. This opens intriguing questions as to potential
mechanisms constraining the variation range of population variability and
synchrony among communities. Lower variation in population synchrony among the
fish communities studied here might relates to their trophic structure as theory
predicts potential synchronising effect of generalist predators, such as
piscivorous fishes, on prey dynamics [@raimondo_interspecific_2004;
@huber_role_2006]. Trophic interactions might also raise synchrony between
predators and preys, [Fig. S9, @bulmer_phase_1975], although predictions can
vary depending on the distribution of interaction strengths [@huber_role_2006;
@mccann_more_2009]. Restricted variations in population synchrony among
communities can also be fostered by low biomass evenness and thereby low
portfolio effect [@doak_statistical_1998; @thibaut_understanding_2013], but
whether biomass eveness of stream fish communities is particularly low remains
to be tested. Alternatively, aquatic ecosystems have been suggested to show
greater population variability than terrestrial ones
[@rip_cross-ecosystem_2011], which could translate in greater variation range of
population variability among aquatic communities than among terrestrial ones.
In any case, more studies are needed to understand how the type of ecosystems
and trophic interactions constrain synchrony and population variability.

@@

sem_bm_caption <- paste0( 
  "Relationships among the predictors of community biomass. (A) Structural
  Equation Model for the median of annual community biomass (in grams by square
    metre). Dotted and solid arrows represent, respectively, negative and
  positive path coefficients. Non significant paths are coloured in grey. The
  path coefficients are standardised. $R^2$: adjusted $R^2$; Avg: Average, CVsp: weighted
  temporal coefficient of variation of the population biomass; PCA1 and PCA2:
  respectively the first and second axis of the principal component analysis
  performed on environmental variables (see methods). (B) to (G): significant
  bivariate relationships between variables of the structural equation model.
  Solid lines depict the predictions of the statistical model. See Fig. S8
  for a sensitivity analysis according to different station selection
  criteria.")

@@

sem_stab_caption <- paste0(
  "Relationships among the predictors of the temporal stability of community
  biomass. (A) Structural Equation Model for the temporal stability of community
  biomass. Dotted and solid arrows represent, respectively, negative and positive
  path coefficients. Non significant paths are coloured in grey. The path
  coefficients are standardised. $R^2$: adjusted $R^2$; Avg: Average, CVsp: weighted
  temporal coefficient of variation of the population biomass; PCA1 and PCA2:
  respectively the first and second axis of the principal component analysis
  performed on environmental variables (see methods). The $R^2$ for the temporal
  stability of community biomass is exactly one because the relation is
  determistic (see Eq. 4). (B) to (I): significant
  bivariate relationships between variables of the structural equation model.
  Solid lines depict the predictions of the statistical model. See Fig. S8
  for a sensitivity analysis according to different station selection
  criteria."
  )
```
